An efficient composition of bidirectional programs by tupling and lazy updates

### 2019-12-02 paper/2.minbigul.tex
change fail to undefined
Polished text

minBiGUL, our target language in this paper, is a subset of BiGUL which is a simple yet powerful putback-based bidirectional language. BiGUL supports two transformations: a forward transformation $get$ producing a view from a source and a backward transformation $put$ taking a source and a modified view to produce an updated source. Intuitively, if we have a BiGUL program $bx$, these two transformations are following functions:

@@

The \textsc{GetPut} law means that if there is no change to the view, there should be no change to the source. The \textsc{PutGet} law means that we can recover the modified view by applying the forward transformation to the updated source.

minBiGUL inherited from BiGUL also supports transformations $put$ and $get$ which satisfy two above laws. In addition, we restrict `adaptive case's of BiGUL on minBiGUL. Then $put$ and $get$ satisfy one more law, \textsc{PutPut} \cite{Foster:2007:CBT:1232420.1232424}, like the following:

@@

The \textsc{PutPut} law means that a source update should overwrite the effect of previous source updates. Due to the satisfaction of three laws, \textsc{GetPut}, \textsc{PutGet} and \textsc{PutPut}, minBiGUL is very well-behaved \cite{Foster:2007:CBT:1232420.1232424}.

@@

\smallvspace
    $bx := Skip \ h \
        | \ Replace \
        | \ Prod \ bx_1 \ bx_2 \
        | \ RearrS \ f_1 \ f_2 \ bx \
        | \ RearrV \ g_1 \ g_2 \ bx \\
    \qtabs{2} | \ Case \ cond_{sv} \ cond_{s} \ bx_1 \ bx_2 \
        | \ Compose \ bx_1 \ bx_2$
\smallvspace

A minBiGUL program may be either a skip of a function or a replacement or a product of two minBiGUL programs or a source/view rearrangement or a case combinator without adaptive cases or a composition of some minBiGUL programs. We use numbers, pairs and lists to construct the program inputs including the source and/or the view.

For source/view rearrangement, BiGUL uses just one lambda expression to express how to deconstruct as well as reconstruct data. It is a kind of bijection. However, to be able to implement it in OCaml, the environment used for developing minBiGUL and solutions in the paper, we need to require two functions which one is the inverse of the other. In the above syntax, $f_2 = f_1^{-1}$ and $g_2 = g_1^{-1}$.

@@

@@

        \noindent $\putbx{Skip \ h}{s}{v} = \\
            \tab \sif{h \ s = v}{s}{\text{fail}}$

@@

We define the semantics of $put$ and $get$ as in Definitions~\ref{def:minbigulput} and \ref{def:minbigulget} respectively. Instead of using $v'$ in the $put$ direction like Figures~\ref{fig:eval-phead}, \ref{fig:eval-comp-phead} and \ref{fig:eval-comp-phead-2}, we just simply use notation $v$ as the updated view. The later definitions also follow this convention.

In the above definitions, we use if-then-else-fi statements to express semantics of $\putbxinline{Case}$ and $\getbxinline{Case}$ where $\Leftarrow$ means assignments. This statement is useful to describe many functions related to $Case$ in this paper. Statement (if $E_1$ then $X_1$ else $X_2$ fi $E_2$) means ``if the test $E_1$ is true, the statement $X_1$ is executed and the assertion $E_2$ must be true, otherwise, if $E_2$ is false, the statement $X_2$ is executed and the assertion $E_2$ must be false.'' If the values of $E_1$ and $E_2$ are distinct, the if-then-else-fi structure is undefined. We can write the equivalent if-then-else statement as follows:

@@

In the end of this section, let us see the minBiGUL definition of $bfoldr$ which is a putback function of an important higher-order function on lists, $foldr$:

\smallvspace
$\tab bfoldr \ bx = Case \ cond_{sv} \ cond_s \ bx_1 \ bx_2$ where:\\
$\tabs{3} \ cond_{sv} = \lambda (s_1,s_2).\lambda v.(s_1 = [\ ]), \ cond_s = \lambda (s_1,s_2).(s_1 = [\ ])$\\
$\tabs{3} \ bx_1 = RearrV \ g_1 \ g_2 \ bx_v$ where: \\
    $\tabs{4} g_1 = g_2^{-1} = \lambda [v].(v,[\ ]), bx_v = (Skip \ (\lambda \_.())) \times Replace$\\
$\tabs{3} \ bx_2 = RearrS \ f_1 \ f_2 \ bx_s$ where:\\
    $\tabs{4} f_1 = f_2^{-1} = \lambda ((x:xs),e).(x,(xs,e))$, 
$bx_s = ((Replace \times bfoldr \ bx) \circ bx)$
\smallvspace

\noindent In $bfoldr$, the composition is inherently left associative, and the number of composition is dynamically determined by the length of the source list. Since $\circ$ has a higher priority than $\times$, it is seemingly impossible to transform $bfoldr$ from the left associative composition style to the right one. 
%will look like a composition of many programs at some points in the evaluation if we slow down this process. Since $\circ$ has a higher priority than $\times$, it is seemingly impossible to transform $bfoldr$ from the left associative composition style to the right one.

>>>>

The target language in this paper, minBiGUL, is a very-well behaved subset of BiGUL, which is a simple, yet powerful putback-based bidirectional language.

BiGUL supports two transformations: a forward transformation $get$ producing a view from a source and a backward transformation $put$ taking a source and a modified view to produce an updated source. Intuitively, if we have a BiGUL program $bx$, these two transformations are the following functions:

@@

\noindent The \textsc{GetPut} law means that if there is no change to the view, there should be no change to the source. The \textsc{PutGet} law means that we can recover the modified view by applying the forward transformation to the updated source.

minBiGUL inherits from BiGUL both 
% supports 
transformations, $put$ and $get$, which satisfy the two laws above. 
% In addition, 
Because we restrict the `adaptive case' of BiGUL in minBiGUL,
% In this case, 
$put$ and $get$ satisfy one more law, namely the \textsc{PutPut} law~\cite{Foster:2007:CBT:1232420.1232424}:
% like the following:

@@

\noindent The \textsc{PutPut} law means that a source update should overwrite the effect of previous source updates. 
Because minBiGUL satisfies all three laws, \textsc{GetPut}, \textsc{PutGet} and \textsc{PutPut}, it is very well-behaved~\cite{Foster:2007:CBT:1232420.1232424}.
% Due to the satisfaction of three laws, \textsc{GetPut}, \textsc{PutGet} and \textsc{PutPut}, minBiGUL is very well-behaved \cite{Foster:2007:CBT:1232420.1232424}.

@@

\smallvspace
    $bx ::= Skip \ h \
        | \ Replace \
        | \ Prod \ bx_1 \ bx_2 \
        | \ RearrS \ f_1 \ f_2 \ bx \
        | \ RearrV \ g_1 \ g_2 \ bx \\
    \qtabs{2} | \ Case \ cond_{sv} \ cond_{s} \ bx_1 \ bx_2 \
        | \ Compose \ bx_1 \ bx_2$
\smallvspace

A minBiGUL program is either a skip of a function, a replacement, a product of two 
%minBiGUL 
programs, a source/view rearrangement, a case combinator (without adaptive cases), or a composition of 
% some minBiGUL 
two programs. We use numbers, pairs and lists to construct the program inputs including the source and/or the view.

For source/view rearrangement, BiGUL uses 
% just one 
a lambda expression to express how to deconstruct as well as reconstruct data. It is a kind of bijection. However, to be able to implement it in OCaml, the environment used for developing minBiGUL and solutions in the paper, we need to require two functions which one is the inverse of the other. In the above syntax, $f_2 = f_1^{-1}$ and $g_2 = g_1^{-1}$.

@@

The semantics of $put$ and $get$ is shown in Definitions~\ref{def:minbigulput} and \ref{def:minbigulget}, respectively. Instead of using the name $v'$ for the updated view in the $put$ direction, like Figures~\ref{fig:eval-phead}, \ref{fig:eval-comp-phead} and \ref{fig:eval-comp-phead-2}, we 
%just
simply 
%use notation 
use~$v$ below. The later definitions also follow this convention.

@@

        \noindent $\putbx{Skip \ h}{s}{v} = \\
            \tab \sif{h \ s = v}{s}{\text{undefined}}$

@@

The two definitions use if-then-else-fi statements 
%express 
to define the semantics of $\putbxinline{Case}$ and $\getbxinline{Case}$, where $\Leftarrow$ means assignments. This statement is useful to describe many functions related to $Case$ in this paper. Statement (if $E_1$ then $X_1$ else $X_2$ fi $E_2$) means ``if the test $E_1$ is true, the statement $X_1$ is executed and the assertion $E_2$ must be true, otherwise, if $E_2$ is false, the statement $X_2$ is executed and the assertion $E_2$ must be false.'' If the values of $E_1$ and $E_2$ are distinct, the if-then-else-fi structure is undefined. We can write the equivalent if-then-else statement as follows:

@@

To close this section,
% In the end of this section, 
let us see the minBiGUL definition of $\textit{bfoldr}$ which is a putback function of an important higher-order function on lists, $\textit{foldr}$:

\smallvspace
$\tab \textit{bfoldr} \ bx = Case \ cond_{sv} \ cond_s \ bx_1 \ bx_2$ where:\\
$\tabs{3} \ cond_{sv} = \lambda (s_1,s_2).\lambda v.(s_1 = [\ ]), \ cond_s = \lambda (s_1,s_2).(s_1 = [\ ])$\\
$\tabs{3} \ bx_1 = RearrV \ g_1 \ g_2 \ bx_v$ where: \\
    $\tabs{4} g_1 = g_2^{-1} = \lambda [v].(v,[\ ]), bx_v = (Skip \ (\lambda \_.())) \times Replace$\\
$\tabs{3} \ bx_2 = RearrS \ f_1 \ f_2 \ bx_s$ where:\\
    $\tabs{4} f_1 = f_2^{-1} = \lambda ((x:xs),e).(x,(xs,e))$, 
$bx_s = ((Replace \times bfoldr \ bx) \circ bx)$
\smallvspace

\noindent In $\textit{bfoldr}$, the composition is inherently left associative, and the number of composition is dynamically determined by the length of the source list. Because~$\circ$ has a higher priority than~$\times$, it is seemingly impossible to transform $\mathit{bfoldr}$ from the left associative composition style to the right one. 
%will look like a composition of many programs at some points in the evaluation if we slow down this process. Since $\circ$ has a higher priority than $\times$, it is seemingly impossible to transform $bfoldr$ from the left associative composition style to the right one.
